---
title: "TfL Oyster Card Analysis"
author: "Graham Cox"
date: "30th May 2022"
output:
  html_document: 
    highlight: espresso
  html_notebook:
    toc: yes
    highlight: espresso
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load libraries 
library(tidyverse)
library(lubridate)
library(showtext)
library(ggtext)
library(tidytext)

# Set up fonts 

font_add_google(family = "patua-one", "Patua One")
font_add_google(family = "roboto-condensed", "Roboto Condensed")

showtext_auto()

# Set main plot theme

theme_set(theme_minimal(base_family = "roboto-condensed") +
            theme(
              plot.title = element_text(family = "patua-one",
                                        size = 22),
              plot.title.position = "plot",
              panel.grid = element_blank()
            ))

```
## A 5% Sample of Oyster Card Journeys on the London Transport Network during 1 week in November 2009.

### Introduction

In this notebook I will analyse a dataset that contains information from TFL Oyster Card journeys. The Oyster card is a contactless payment method and was first introduced in London in June 2003. The card can contain pre-paid tickets for a period of time, travel permits and pre-paid credit for travel, commonly known as Pay-As-You-Go, or PAYG.

The card allows for several on the variety of transport modes available to passengers

- London Underground
- London Buses
- Nation Rail Overground Network
- Docklands Light Railway
- London Tramlink

The Oyster card can also be used on the Emirates Skyline cable car and on certain river boat services. **Data for journeys on these transport modes is not included in this dataset.**

### Questions

1. Does the definition of "rush hour" of the majority of journeys being made between 08:00 and 09:00 and 17:00 and 18:00 hold true with this dataset?
2. What are the busiest locations?
3. What are the top 5 busiest locations within each London Underground Zones 1-9?
4. What are the busiest times of day for each type of Oyster Card?
5. What is the average amount spent for journeys during the rush hour?

### Load Data and Initial Wrangling

```{r data-load}

# Initial load of data

df <- read_csv("../data/Nov09JnyExport.csv",
               show_col_types = FALSE)

# View dataset

glimpse(df)

```

Let's clean up the variable names using the `janitor` package.

```{r data-clean-1}

df <- df %>% 
  janitor::clean_names()

glimpse(df)
```

#### Variable Review

Reviewing the variables available in the dataset, the following assumptions are made

- Where `start_stn` equals "Unstarted" or `end_station` equals "Unfinished" these are journeys where the passenger failed to tap-in or tap-out with an Oyster Card at the start or end of their journeys. These observations also have a value of 00:00:00 in either the `ent_time_hhmm` or `ex_time_hhmm` variables. **These observations will be removed**
- The `ZVPPT` variable does not contain meaningful information. **This variable will be removed**
- The `ent_time` and `ex_time` variables do not hold meaningful information. **These variables will be removed**


```{r}
df <- df %>% 
  select(-c(ent_time, ex_time, zvppt)) %>% 
  filter(!if_any(start_stn:end_station, ~ .x %in% c("Unstarted", "Unfinished")))
```
---  
  
#### Review of the `sub_system` variable
  
```{r}
df %>% select(sub_system) %>% group_by(sub_system) %>%
  ggplot(aes(sub_system)) +
  geom_bar() +
  coord_flip()
```

- The highest number of observations are for those journeys on LTB. These observations have a `start_stn` and `end_station` of "BUS". These observations also have an ex_time_hhmm of "00:00:00". As passengers are not required to tap-out at the end of the journey, an end time cannot be recorded. **These observations will be removed**
- The remaining `sub_systems` have a considerably lower number of observations compared to the new highest group of "LUL" i.e. London Underground. **All other `sub_system` values will now be removed**
- As the variable named `sub_system` now only contains one value of "LUL". **This variable can now be removed**
- The variable named `route_id` was for bus journeys only. **This variable can now be removed**

```{r}
final_df <- df %>% 
  filter(sub_system == "LUL") %>% 
  select(-c(sub_system, route_id))

glimpse(final_df)

```